---
title: "Project02 - Group01"
author: "Eva, Tobi, Kathi, Laura"
date: "14 Juni 2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library packages
library(cluster) 
library(dendextend)
library(dplyr)
library(factoextra)
library(fpc)
library(ggplot2)
library(ggpubr)
library(matrixTests)
library(pheatmap)
library(RColorBrewer)
library(viridis)
library(plotly)
library(ggrepel)
library(compare)
library(data.table)
library(matrixTests)

```


## data loading
```{r }


wd = getwd()

NCI_TPW_gep_treated = readRDS(paste0(wd, "/Data/NCI_TPW_gep_treated.rds"))
NCI_TPW_gep_untreated = readRDS(paste0(wd, "/Data/NCI_TPW_gep_untreated.rds"))
Metadata = read.delim(paste0(wd, "/Data/NCI_TPW_metadata.tsv"), header = TRUE, sep = "\t", stringsAsFactors = TRUE)
Cellline_Annotation = read.delim(paste0(wd, "/Data/cellline_annotation.tsv"), header = TRUE, sep = "\t", stringsAsFactors = TRUE)
Drug_Annotation = read.delim(paste0(wd, "/Data/drug_annotation.tsv"), header = TRUE, sep = "\t", stringsAsFactors = TRUE)
CCLE_mutations = readRDS(paste0(wd, "/Data/CCLE_mutations.rds"))
CCLE_copynumber = readRDS(paste0(wd, "/Data/CCLE_copynumber.rds"))
CCLE_basalexpression = readRDS(paste0(wd, "/Data/CCLE_basalexpression.rds"))
NegLogGI50 = as.data.frame(readRDS(paste0(wd, "/Data/NegLogGI50.rds")))
Treated = data.frame(NCI_TPW_gep_treated)
Untreated = data.frame(NCI_TPW_gep_untreated)
```

## data scaling
After checking for normalization, we scaled our data in the first place to provide the scaled data for further analysis.
```{r }
list = list(Treated,Untreated)
nlist = lapply(list,scale)
Treated = as.data.frame(nlist[[1]])
Untreated = as.data.frame(nlist[[2]])
Fold_Change = Treated - Untreated
Fold_Change = data.frame(Fold_Change)
rm(NCI_TPW_gep_treated,NCI_TPW_gep_untreated,list,nlist)

```

# 1. Broad analysis

###Boxplots (already normalized)
This step was done before scaling the data. The boxplots showed a deviation which is the reason for scaling the data.
```{r}
boxplot(Treated)

boxplot(Treated,  ylab = "Gene expression profile", main = "Treated genexpressionprofiles",xaxt = "n")

boxplot(Untreated, ylab = "Gene expression profile", main = "Untreated genexpressionprofiles",xaxt = "n")

boxplot(Treated[,1:10], ylab = "Gene expression profile", main = "First 10 reated genexpressionprofiles")
```


```{r coloured boxplots, creating a susbet of Metadata based on the amount of tested genes in Treated}
Treated1 = readRDS(paste0(wd, "/Data/NCI_TPW_gep_treated.rds"))
df = data.frame(t(Treated1))
df.data <- data.frame(sample = rownames(df))
adjustedMeda = subset(Metadata, sample %in% intersect(Metadata$sample, df.data$sample))
rm(df,df.data, Treated1)
palette(rainbow(15))
boxplot(Treated, border=adjustedMeda$drug,xlab= "Different Drugs" ,ylab = "Gene expression profile", main = "Teated genexpressionprofiles",xaxt ="n")
```

### Densityplot 
The abline shows the 3 quantiles (      25%       50%       75%    )
```{r}
NCI_TPW_gep_treated = readRDS(paste0(wd, "/Data/NCI_TPW_gep_treated.rds"))
NCI_TPW_gep_untreated = readRDS(paste0(wd, "/Data/NCI_TPW_gep_untreated.rds"))
plot(density(NCI_TPW_gep_treated), "Densityplot Treated vs Untreated")
lines(density(NCI_TPW_gep_untreated), col = "indianred2")
legend("topright", legend = c("treated", "untreated"), col = c("black", "indianred2"), pch = 20)
abline(v = quantile(NCI_TPW_gep_treated)[2:4], col = c("lightblue", "blue",  "orange"), lty = 2)
```


###k-means clustering
To look for clusters in the raw data we performed a k-menas clustering and searched for potentially clusters.
```{r}
# Performing a k-means on Treated
#Determining the number of clusters
topVarTreated = apply(Treated, 1, var)
summary(topVarTreated)

# Using the most variable, thus informative genes
topVarTreated75 = Treated[topVarTreated > quantile(topVarTreated, probs = 0.75), ]
dim(topVarTreated75)

```

```{r}
km = kmeans(x = t(topVarTreated75), centers = 3, nstart = 10)
km$tot.withinss
```

```{r}
km = kmeans(x = t(topVarTreated75), centers = 2, nstart = 10)
km$tot.withinss
```

```{r}
#running a loop for the best n (searching for "ellbow")
wss = sapply(2:7, function(k) {
kmeans(x = t(topVarTreated75), centers = k)$tot.withinss})            
plot(2:7, wss, type = "b", pch = 19, xlab = "Number of clusters K", ylab = "Total within-clusters sum of squares", main = "Determining the amount of clusters from Treated")
```
As we wanted an "ellbow"to get a good result we can say in a way that our data are not really good to cluster.
To look in a other way, we also provided the clusters by the silhouette-method.
```{r,echo=FALSE}
# Using the silhouett method
D = dist(t(topVarTreated75))
km = kmeans(x = t(topVarTreated75), centers = 10, nstart = 10)
s = silhouette(km$cluster, D)
plot(s)
```


###PCA
```{r}
pca <- prcomp(t(Fold_Change), scale = TRUE)

```
```{r}
# sdev calculates variation each PC accounts for
pca.var <- pca$sdev^2  
# since percentages make more sense then normal variation values
# calculate % or variation, which is much more interesing
pca.var.per <- round(pca.var/sum(pca.var)*100, 1) 

barplot(pca.var.per, main = "Scree plot", xlab = "Principal Components", ylab = "% variation")
plot(pca.var.per[1:10], main = "Elbow plot", type = "l", xlab = "Principal Components", ylab = "% variation")
plot(cumsum(pca.var.per[1:15]), main = "cumulative variation", type = "l", xlab = "Principal Components", ylab = "% variation") 

```
```{r}
#creating data frame with all pcs 
#cleaning up sample names as they differed between matrices 
pca.data <- data.frame(pca$x)
rownames(pca.data) <- gsub(x = rownames(pca.data), pattern = "X786", replacement = "786")
pca.data <- cbind(sample =rownames(pca.data), pca.data)

```
```{r}
## get names of top 10 genes that contribute most to pc1
loading_scores_1 <- pca$rotation[,1]
gene_score <- abs(loading_scores_1) ## sort magnitude
gene_score_ranked <- sort(gene_score, decreasing = TRUE)


top_10_genes <- names(gene_score_ranked[1:10])
top_10_genes # show names of top 10 genes
```
```{r}
### Metadata color matrix for coloring 
Metadata$sample <- gsub(x = Metadata$sample, pattern = "-", replacement = ".")

metad.cl <- subset(Metadata, Metadata$sample %in% pca.data$sample) 
## adjust row length of metadata to pca.data


metad.cl$mechanism <- Drug_Annotation$Mechanism[match(metad.cl$drug, Drug_Annotation$Drug)]
metad.cl$msi <- Cellline_Annotation$Microsatellite_instability_status[match(metad.cl$cell, Cellline_Annotation$Cell_Line_Name)]

```
```{r}

# plotting all informative PCs
#color vectors for coloring by drug and tissue
viridis <- viridis(9)
color_tissue = viridis[metad.cl$tissue]
tissue <- levels(metad.cl$tissue)

magma <- magma(15)
color_drug = magma[metad.cl$drug]
drug <- levels(metad.cl$drug)


## colored by drug
#plot PC1 and PC2
plot(pca$x[,1], 
     pca$x[,2], 
     col = color_drug,
     pch = 19, 
     xlab = paste("PC1 (",pca.var.per[1],"%)"), 
     ylab = paste("PC2 (",pca.var.per[2],"%)"))
#create legend
legend("topleft", 
       legend = drug, 
       col = magma, 
       pch = 19, 
       xpd = "TRUE",
       bty = "n",
       cex = 0.75
)
#create title
mtext("PCA of Fold Change  colored by drug", 
      side = 3, 
      line = -2,
      cex = 1.2,
      font = 2, 
      outer = TRUE)

#plot PC2 and PC3
plot(pca$x[,2], 
     pca$x[,3], 
     col = color_drug,
     pch = 19, 
     xlab = paste("PC2 (",pca.var.per[2],"%)"), 
     ylab = paste("PC3 (",pca.var.per[3],"%)"))
#create legend
legend("right", 
       legend = drug, 
       col = magma, 
       pch = 19, 
       xpd = "TRUE",
       bty = "n",
       cex = 0.75,
       inset = c(0, 2)
)
#create title
mtext("PCA of Fold Change  colored by drug", 
      side = 3, 
      line = -2,
      cex = 1.2,
      font = 2, 
      outer = TRUE)




## colored by tissue
#plot PC1 and PC2
plot(pca$x[,1], 
     pca$x[,2], 
     col = color_tissue,
     pch = 19, 
     xlab = paste("PC1 (",pca.var.per[1],"%)"), 
     ylab = paste("PC2 (",pca.var.per[2],"%)"))
#create legend
legend("topleft", 
       legend = tissue, 
       col = viridis, 
       pch = 19, 
       xpd = "TRUE",
       bty = "n",
       cex = 0.75
)
#create title
mtext("PCA of Fold Change  colored by tissue", 
      side = 3, 
      line = -2,
      cex = 1.2,
      font = 2, 
      outer = TRUE)

#plot PC2 and PC3
plot(pca$x[,2], 
     pca$x[,3], 
     col = color_tissue,
     pch = 19, 
     xlab = paste("PC2 (",pca.var.per[2],"%)"), 
     ylab = paste("PC3 (",pca.var.per[3],"%)"))
#create legend
legend("right", 
       legend = tissue, 
       col = viridis, 
       pch = 19, 
       xpd = "TRUE",
       bty = "n",
       cex = 0.75,
       inset = c(0, 2)
)
#create title
mtext("PCA of Fold Change  colored by tissue", 
      side = 3, 
      line = -2,
      cex = 1.2,
      font = 2,
      outer = TRUE)

```
