---
title: "Project02 - Group01"
author: "Eva, Tobi, Kathi, Laura"
date: "24. July 2019"
output: html_document
 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      chache = TRUE)
#library packages
library(cluster) 
library(compare)
library(data.table)
library(dendextend)
library(dplyr)
library(factoextra)
library(fpc)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(matrixTests)
library(pheatmap)
library(plotly)
library(RColorBrewer)
library(viridis)

```

# Table of content
[4. Main questions](#anchor4)  
  [4.1 Question 1: Does the cell doubling time correlate with reduced drug sensitivity?](#css_id4.1)  
    [4.1.1 Linear regression ~ Foldchange-means](#css_class4.1.1)  
    [4.1.2 Linear regression ~ Foldchange-means biomarkers](#css_class4.1.2)  
    [4.1.3 Linear regression ~ Doubling-time](#css_class4.1.3)  
    [4.1.4 Multiple regression](#css_class4.1.4)  
    [4.1.5 Table of conclusion](#css_class4.1.5)  

  [4.2 Question 2: Does Lapatinib have a similar effect on lung cancer as Erlotinib?](#css_id4.2)  
    [4.2.1 Heatmap](#css_class4.2.1)  
    [4.2.2 Plot Erlotinib and Lapatinib, coloured by tissue](#css_class4.2.2)  
    [4.2.3 Pearson correlation](#css_class4.2.3)  
    [4.2.4 Lung cellines](#css_class4.2.4)  
    [4.2.5 Anova](#css_class4.2.5)  

  [4.3 Question 3: Comparing lapatinib treated breast and cns celllines](#css_id4.3)  
    [4.3.x Volcano plot](#css_class4.3.x)
  
  

```{r, echo=FALSE}


wd = getwd()

NCI_TPW_gep_treated = readRDS(paste0(wd, "/Data/NCI_TPW_gep_treated.rds"))
NCI_TPW_gep_untreated = readRDS(paste0(wd, "/Data/NCI_TPW_gep_untreated.rds"))
Metadata = read.delim(paste0(wd, "/Data/NCI_TPW_metadata.tsv"), header = TRUE, sep = "\t", stringsAsFactors = TRUE)
Cellline_Annotation = read.delim(paste0(wd, "/Data/cellline_annotation.tsv"), header = TRUE, sep = "\t", stringsAsFactors = TRUE)
Drug_Annotation = read.delim(paste0(wd, "/Data/drug_annotation.tsv"), header = TRUE, sep = "\t", stringsAsFactors = TRUE)
CCLE_mutations = readRDS(paste0(wd, "/Data/CCLE_mutations.rds"))
CCLE_copynumber = readRDS(paste0(wd, "/Data/CCLE_copynumber.rds"))
CCLE_basalexpression = readRDS(paste0(wd, "/Data/CCLE_basalexpression.rds"))
NegLogGI50 = as.data.frame(readRDS(paste0(wd, "/Data/NegLogGI50.rds")))
Treated = data.frame(NCI_TPW_gep_treated)
Untreated = data.frame(NCI_TPW_gep_untreated)
```

```{r ,echo=FALSE}
list = list(Treated,Untreated)
nlist = lapply(list,scale)
Treated = as.data.frame(nlist[[1]])
Untreated = as.data.frame(nlist[[2]])
Fold_Change = Treated - Untreated
Fold_Change = data.frame(Fold_Change)
rm(NCI_TPW_gep_treated,NCI_TPW_gep_untreated,list,nlist)

```

# 4. Main questions {#anchor4}


## 4.1 Question 1: Does the cell doubling time correlate with reduced drug sensitivity? {#css_id4.1}

<div style = "text-align: justify">
As mentioned in our presentation, we want to create a model to predict GI50-values and thus to predict, if Lapatinib is a good choice. Therefore, we took several datasets and performed 4 regression analyses.


### 4.1.1 Linear regression ~ Foldchange-means {#css_class4.1.1}

The first linear model tries to predict the G-50 value under the data of the Foldchange-means.


```{r, echo = FALSE}

NegLogGI50Lap = NegLogGI50[9,]
Fold_ChangeLap = select(Fold_Change, contains("Lapa"))
Fold_ChangeLap_t = t(Fold_ChangeLap)
means = colMeans(Fold_ChangeLap)
Fold_Changemeans = as.data.frame(t(means))

a2 = gsub(x = colnames (Fold_Changemeans), pattern = "_lapatinib_10000nM_24h", replacement = "")
colnames(Fold_Changemeans) = a2

a3 = gsub(x = a2, pattern = "X7", replacement = "7")
colnames(Fold_Changemeans) = a3


a1 = gsub(x = colnames (NegLogGI50Lap), pattern = "-", replacement = ".")
colnames(NegLogGI50Lap) = a1

c1 = rbind(a1,NegLogGI50Lap)
c2 = rbind(a3,Fold_Changemeans)

c1 = t(c1)
c2 = t(c2)

c1 =as.data.frame(c1)
c2 =as.data.frame(c2)


c3 = subset(c1, `1` %in% intersect(c1$`1`, c2$V1))
c4 = as.numeric(as.character(c3$lapatinib))
adjustedNeglogGI50Lap = as.data.frame(c4)



Fold_Changemeans = as.data.frame(t(Fold_Changemeans))

combined1 = cbind(adjustedNeglogGI50Lap, Fold_Changemeans)

names1 = c( "NegLogGI50Lap","Fold_Changemeans")
colnames(combined1) = names1
                      
lmFold = lm(NegLogGI50Lap ~ Fold_Changemeans, data = combined1)

#cor(combined1$NegLogI50Lap,combined1$Fold_Changemeans)

n = nrow(combined1)
rmse1 = sqrt(1/n * sum(lmFold$residuals^2))
#rmse1

#Split the data (Training - Testing)

i1.train = sample(1:nrow(combined1), 44)

dat1.train = combined1[i1.train, ]
dat1.test = combined1[-i1.train, ]

l1.train = lm(NegLogGI50Lap ~ Fold_Changemeans, data = dat1.train)
#summary(l1.train)

qqnorm(l1.train$residuals, main = "Test for normaldistribution of residuals")
qqline(l1.train$residuals)

plot(dat1.train$NegLogGI50Lap, l1.train$fitted.values, pch = 20, col = "blue", xlab = "Real values", 
     ylab = "Predicted values", main = "Linear regression ~ Fold_Changemeans ~ Training")
abline(0, 1, col = "red")

n = nrow(dat1.train)
rmse1.train = sqrt(1/n * sum(l1.train$residuals^2))
#rmse1.train

pred1 = predict(l1.train, newdata = dat1.test)

n = nrow(dat1.test)
residuals = dat1.test$NegLogGI50Lap - pred1
rmse1.test = sqrt(1/n * sum(residuals^2))
#rmse1.test

```

### 4.1.2 Linear regression ~ Foldchange-means biomarkers {#css_class4.1.2}

Linear regression is not suitable here, as the mean fold-change per cell line is not a meaningful indicator in a biological context.
Hence, we selected the top 20 genes to calculate our mean values. Afterwards linear regression was performed.

```{r, echo = FALSE}
Treated_t<-data.frame(t(Treated))
Untreated_t<-data.frame(t(Untreated))
LapatinibUntreated<-select(Untreated, contains("Lapa"))
LapatinibTreated<-select(Treated, contains("Lapa"))
LapTreat=as.data.frame(t(LapatinibUntreated))
LapUn=as.data.frame(t(LapatinibTreated))

m1<-colMeans(LapTreat, na.rm = TRUE)
m2<-colMeans(LapUn, na.rm = TRUE)

ut<-(m1-m2)
utabs = abs(ut)
ut_sort<-sort(utabs, decreasing = TRUE)
biomark= as.data.frame(ut_sort[1:20])
n1 = rownames(biomark)
biomark = cbind(n1,biomark)


Fold_ChangeLap = select(Fold_Change, contains("Lapa"))
Fold_ChangeLap_t = as.data.frame(Fold_ChangeLap)
Fold_ChangeLapBM = subset(Fold_ChangeLap_t, rownames(Fold_ChangeLap_t) %in% biomark$n1)

NegLogGI50Lap = NegLogGI50[9,]
means = colMeans(Fold_ChangeLapBM)
Fold_ChangemeansBM = as.data.frame(t(means))

a2 = gsub(x = colnames (Fold_ChangemeansBM), pattern = "_lapatinib_10000nM_24h", replacement = "")
colnames(Fold_ChangemeansBM) = a2

a3 = gsub(x = a2, pattern = "X7", replacement = "7")
colnames(Fold_ChangemeansBM) = a3


a1 = gsub(x = colnames (NegLogGI50Lap), pattern = "-", replacement = ".")
colnames(NegLogGI50Lap) = a1

c1 = rbind(a1,NegLogGI50Lap)
c2 = rbind(a3,Fold_ChangemeansBM)

c1 = t(c1)
c2 = t(c2)

c1 =as.data.frame(c1)
c2 =as.data.frame(c2)


c3 = subset(c1, `1` %in% intersect(c1$`1`, c2$V1))
c4 = as.numeric(as.character(c3$lapatinib))
adjustedNeglogGI50Lap = as.data.frame(c4)



Fold_ChangemeansBM = as.data.frame(t(Fold_ChangemeansBM))

combined11 = cbind(adjustedNeglogGI50Lap, Fold_ChangemeansBM)

names11 = c( "NegLogGI50Lap","Fold_ChangemeansBM")
colnames(combined11) = names11
                      
lmFold1 = lm(NegLogGI50Lap ~ Fold_ChangemeansBM, data = combined11)

#cor(combined1$NegLogI50Lap,combined1$Fold_Changemeans)

n = nrow(combined11)
rmse11 = sqrt(1/n * sum(lmFold1$residuals^2))
#rmse11

#Split the data (Training - Testing)

i11.train = sample(1:nrow(combined11), 44)

dat11.train = combined11[i11.train, ]
dat11.test = combined11[-i11.train, ]

l11.train = lm(NegLogGI50Lap ~ Fold_ChangemeansBM, data = dat11.train)
#summary(l11.train)

qqnorm(l11.train$residuals, main = "Test for normaldistribution of residuals")
qqline(l11.train$residuals)

plot(dat11.train$NegLogGI50Lap, l11.train$fitted.values, pch = 20, col = "blue", xlab = "Real values", 
     ylab = "Predicted values", main = "Linear regression ~ Fold_ChangemeansBM ~ Training")
abline(0, 1, col = "red")

n = nrow(dat11.train)
rmse11.train = sqrt(1/n * sum(l11.train$residuals^2))
#rmse11.train

pred11 = predict(l11.train, newdata = dat11.test)

n = nrow(dat11.test)
residuals = dat11.test$NegLogGI50Lap - pred11
rmse11.test = sqrt(1/n * sum(residuals^2))
#rmse11.test
```


### 4.1.3 Linear regression ~ Doubling-time {#css_class4.1.3}

The second linear model tries to predict the G-50 value under the data of the Doubling-time.
```{r,echo = FALSE}
NegLogGI50Lap = NegLogGI50[9,]

#Sort by Cellline-Name
df = arrange(Cellline_Annotation, Cell_Line_Name)
Doublingtime = cbind.data.frame (df$Cell_Line_Name, df$Doubling_Time)

c21 = as.data.frame(t(NegLogGI50Lap))

combined2 = cbind(c21, Doublingtime$`df$Doubling_Time`)
names2 = c( "NegLogGI50Lap","Doubling_Time")
colnames(combined2) = names2

combined2 =na.omit(combined2)

lmDouble = lm(NegLogGI50Lap ~ Doubling_Time, data = combined2)

#summary(lmDouble)

#cor(combined2$NegLogI50Lap,combined2$Doubling_Time)



#Split the data (Training - Testing)

n = nrow(combined2)
rmse2 = sqrt(1/n * sum(lmDouble$residuals^2))
#rmse2

i2.train = sample(1:nrow(combined2), 48)

dat2.train = combined2[i2.train, ]
dat2.test = combined2[-i2.train, ]

l2.train = lm(NegLogGI50Lap ~ Doubling_Time, data = dat2.train)
#summary(l2.train)

qqnorm(l2.train$residuals, main = "Test for normaldistribution of residuals")
qqline(l2.train$residuals)

plot(dat2.train$NegLogGI50Lap, l2.train$fitted.values, pch = 20, col = "blue", xlab = "Real values", 
     ylab = "Predicted values", main = "Linear regression ~ Doubling-Time ~ Training")
abline(0, 1, col = "red")

n = nrow(dat2.train)
rmse2.train = sqrt(1/n * sum(l2.train$residuals^2))
#rmse2.train

pred2 = predict(l2.train, newdata = dat2.test)

n = nrow(dat1.test)
residuals = dat2.test$NegLogGI50Lap - pred2
rmse2.test = sqrt(1/n * sum(residuals^2))
#rmse2.test
```


### 4.1.4 Multiple regression {#css_class4.1.4}

Finally, we did a multiple regression with both datasets to predict GI50-values.
```{r,echo = FALSE}
b1 = gsub(x =Doublingtime$`df$Cell_Line_Name`, pattern = "-", replacement = ".")
Doublingtime1 =  rbind(b1,Doublingtime$`df$Doubling_Time`)
Doublingtime1 = as.data.frame(t(Doublingtime1)) 

c31 = subset(Doublingtime1, b1 %in% intersect(Doublingtime1$b1, c2$V1))
c41 = as.numeric(as.character(c31$V2))
adjustedDoubling_Time = as.data.frame(c41)

combined3 = cbind(adjustedNeglogGI50Lap, Fold_ChangemeansBM, adjustedDoubling_Time)
names3 = c( "NegLogGI50Lap","Fold_ChangemeansBM","Doubling_Time")
colnames(combined3) = names3

mlr = lm(NegLogGI50Lap ~ ., data = combined3)

#summary(mlr)


#Split the data (Training - Testing)

n = nrow(combined3)
rmse3 = sqrt(1/n * sum(mlr$residuals^2))
#rmse3

i3.train = sample(1:nrow(combined3), 44)

dat3.train = combined3[i3.train, ]
dat3.test = combined3[-i3.train, ]

l3.train = lm(NegLogGI50Lap ~ ., data = dat3.train)
#summary(l3.train)

qqnorm(l3.train$residuals, main = "Test for normaldistribution of residuals")
qqline(l3.train$residuals)

plot(dat3.train$NegLogGI50Lap, l3.train$fitted.values, pch = 20, col = "blue", xlab = "Real values", 
     ylab = "Predicted values", main = "Multiple regression ~ Training")
abline(0, 1, col = "red")

n = nrow(dat3.train)
rmse3.train = sqrt(1/n * sum(l3.train$residuals^2))
#rmse3.train

pred3 = predict(l3.train, newdata = dat3.test)

n = nrow(dat3.test)
residuals = dat3.test$NegLogGI50Lap - pred3
rmse3.test = sqrt(1/n * sum(residuals^2))
#rmse3.test
```


### 4.1.5 Table of conclusion {#css_class4.1.5}

The following table shows important values. 

```{r, results='asis',echo = FALSE}
Rsrd =  c(summary(l1.train)$r.squared, summary(l11.train)$r.squared, summary(l2.train)$r.squared, summary(l3.train)$r.squared)
F_Stat = c(summary(l1.train)$fstatistic[1],summary(l11.train)$fstatistic[1],
           summary(l2.train)$fstatistic[1],summary(l3.train)$fstatistic[1])
Sum_RMSE = c(rmse1.test,rmse11.test,rmse2.test,rmse3.test)
Rsrd1 = as.data.frame(t(Rsrd))
table = rbind(Rsrd1,F_Stat,Sum_RMSE)
colnames(table) = c("Foldchange","FoldchangeBM","Doublingtime","Multiple regression")
rownames(table) = c("R-squared-value","F-statistic","RMSE of test-dataset")
knitr::kable(table, format = "markdown")

```


The data is hard to discuss, because we used random values for our test and training dataset. Hence, different values on every run are obtained. We could use "selected values" (First 20) but this would be kind of cheating.


## 4.2 Question 2: Does Lapatinib have a similar effect on lung cancer as Erlotinib? {#css_id4.2}

As we read in the paper "Antitumor and antiangiogenic effect of the dual EGFR and HER-2 tyrosine kinase inhibitor lapatinib in a lung cancer model. (Diaz et al., 2010)", lapatinib and erlotinib are said to have similar effects. To verify this, we first looked at the correlation of the drugs and then created a graph showing the difference between GI50 for a certain cell line and the middle GI50 for both drugs.

```{r correlation, echo = FALSE}
# correlation in general 
n= as.data.frame(t(NegLogGI50))
rmv.rows = apply(n, 1, function(x) {
  sum(is.na(x))
})
NLGI50.all = n[-which(rmv.rows > 0), ]  # Removing any row with 1 or more missing values
rm(rmv.rows, n, NegLogGI50)

```


### 4.2.1 Heatmap {#css_class4.2.1}
```{r correlation heatmap, echo = FALSE}
cor = cor(NLGI50.all)
pheatmap(cor, col = cm.colors(256), main = "Correlation NLGI50")

```

Erlotinib and lapatinib have a strong correlation.


### 4.2.2 Plot Erlotinib and Lapatinib, coloured by tissue {#css_class4.2.2}
```{r plot erlotinib, fig.height=12, fig.width=16, echo = FALSE}

#differece
diff = data.frame(erlotinib = NLGI50.all$erlotinib - mean(NLGI50.all$erlotinib), lapatinib = NLGI50.all$lapatinib- mean(NLGI50.all$lapatinib))
diff$celllines = rownames(NLGI50.all)
#create vector to insert column tissue from Metadata

tissue = sapply(1:nrow(diff), function(x) {
  position = which(as.character(Metadata$cell) == diff[x, "celllines"])[1] #if tissue occurs several times, take the first
  out = as.character(Metadata[position, "tissue"]) #output the tissue at this position
  return(out)
})
diff$tissue = tissue
rm(tissue)

diff$celllines = factor(diff$celllines, levels = diff$celllines[order(diff$tissue)]) #Classified by tissue

e =  ggplot(diff, aes(x = celllines, y = erlotinib, fill = tissue))+ 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  labs(title = "Mean graph plot of NLGI50 values for Erlotinib")

l = ggplot(diff, aes(x = celllines, y = lapatinib, fill = tissue)) + 
  geom_bar(stat="identity") + 
  coord_flip() + 
  labs(title="Mean graph plot of NLGI50 values for Lapatinib")

ggarrange(e, l,
          labels = c("A", "B"),
          ncol = 2, nrow = 1)
rm(e,l)

```

The difference from the NegLogGI50 for a particular cell line and the mean NegLogGI50 is plotted here for Erlotinib and Lapatinib.

### 4.2.3 Pearson correlation {#css_class4.2.3}
```{r correlation erlotinib, lapatinib, echo = FALSE}
cor(NLGI50.all$erlotinib, NLGI50.all$lapatinib, method = "pearson")
```

A Pearson correlation coefficient of ~ 0.65 confirms that these patterns are similar. One reason for this is that both erlotinib and lapatinib are EGFR inhibitors. Cell lines that were more sensitive are displayed as bars that project to the right of the mean. Cell lines that were less sensitive are displayed with bars projected to the left.

### 4.2.4 Lung cellines {#css_class4.2.4}
To answer our more specific question, whether lapatinib also has an effect on lung cancer, we will only look at the cell lines in lung cancer.

```{r lung genes, fig.height = 7, fig.width = 14, echo = FALSE}
#only lung with mean all
### load data
Metadata_Lapatinib_treated = Metadata[which(Metadata$drug == "lapatinib" & Metadata$dose != "0nM"),]
NegLogGI50 = as.data.frame(readRDS(paste0(wd, "/Data/NegLogGI50.rds")))
#lung cells from Metadata
Lung_Metadata_L_treated = Metadata[which(Metadata$drug == "lapatinib" & Metadata$dose != "0nM" & Metadata$tissue == "Lung"),] 
celllines = Lung_Metadata_L_treated$cell
NegLogGI50.lung = as.data.frame(t(NegLogGI50[c("erlotinib", "lapatinib"), celllines]))

#Difference
dif.NegLogGI50.lung = data.frame(erlotinib = NegLogGI50.lung$erlotinib -  mean(NLGI50.all$erlotinib), lapatinib = NegLogGI50.lung$lapatinib -  mean(NLGI50.all$lapatinib)) #erlotinib data - mean value, lapatinib data - mean value
dif.NegLogGI50.lung$celllines = rownames(NegLogGI50.lung)

# PLot Erlotinib and Lapatinib

e = ggplot(dif.NegLogGI50.lung,aes(x = celllines, y = erlotinib)) + geom_bar(stat = "identity", fill = "skyblue") + geom_text(aes(label=round(erlotinib, 2)), vjust = -0.5, color = "black", size = 3) + coord_flip() + labs(title = "Mean graph plot of NLGI50 values for Erlotinib, only lung genes")
l = ggplot(dif.NegLogGI50.lung,aes(x = celllines, y = lapatinib)) + geom_bar(stat = "identity", fill = "skyblue") + geom_text(aes(label=round(lapatinib, 2)), vjust = -0.5, color = "black", size = 3) + coord_flip() + labs(title = "Mean graph plot of NLGI50 values for Lapatinib, only lung genes")

ggarrange(e, l,
          labels = c("A", "B"),
          ncol = 2, nrow = 1)
rm(e,l)

```

The difference from the NegLogGI50 for cell line from lung tissue and the mean NegLogGI50 is plotted here for Erlotinib and Lapatinib.


```{r pearson correlation lung, echo = FALSE}
#correlation lung

cor(NegLogGI50.lung$erlotinib, NegLogGI50.lung$lapatinib, method = "pearson")

```

A pearson correlation coefficent of ~ 0.96 suggests that Lapatinib has a similar effect on lung cancer as Erlotinib.

<div style="text-align: justify"> 

####Selection of Lapatinib and Erlotinib treated cells
</div>
```{r, echo = FALSE}
lapa<-data.frame(Metadata[which(Metadata[,'drug'] == "lapatinib"), ])
erlo<-data.frame(Metadata[which(Metadata[,'drug'] == "erlotinib"), ])
el<-right_join(lapa,erlo, by="cell")
rmv.rows = apply(el, 1, function(x) {
  sum(is.na(x))
})  # Go through each row and sum up all missing values
row.names(rmv.rows)
```

<div style="text-align: justify"> 
Create data frame with Lapatinib and Erlotinib data </div>
```{r, echo = FALSE}
fc<-data.frame(Treated-Untreated)
all<-data.frame(fc[grep("lapatinib|erlotinib", colnames(fc))])


all.rmv<-data.frame(all %>% dplyr::select(
                    -"CCRF.CEM_erlotinib_10000nM_24h", 
                    -"HL.60_erlotinib_10000nM_24h", 
                    -"HT29_erlotinib_10000nM_24h", 
                    -"K.562_erlotinib_10000nM_24h", 
                    -"LOX_erlotinib_10000nM_24h",
                    -"SR_erlotinib_10000nM_24h",
                    -"COLO.205_lapatinib_10000nM_24h"))
                  
la<-data.frame(all.rmv[grep("lapatinib", colnames(all.rmv))])

er<-data.frame(all.rmv[grep("erlotinib", colnames(all.rmv))])
erla<-data.frame(er,la)
drug<-c(rep('Erlotinib',53), rep('Lapatinib',53))
fc_drug<-apply(erla, MARGIN = 2, FUN = mean)
df_drug<-data.frame(fc_drug, drug)

```

<div style="text-align: justify"> 

### 4.2.5 Anova {#css_class4.2.5}

First, the fold change data of lapatinib and erlotinib must be checked for normality. Only when the data is normalized can anova be used.</div>

```{r}
qqnorm(fc_drug, pch = 1, frame = FALSE)
qqline(fc_drug, col = "steelblue", lwd = 2)
```
<div style="text-align: justify"> 
Anova anticipates variance homogeneity. If the p-value of this test is greater than 0.05, the variance of variance is not rejected and this condition is met.</div>
```{r}
fligner.test(fc_drug ~ drug, data=df_drug )
```

p = 0.53 means that the result is not significant. Thus, the two drugs did not differ significantly from each other.

```{r}
ggboxplot (data = df_drug, x="drug", y="fc_drug", color = "drug",    
          add = "jitter", legend = "none")+
          rotate_x_text(angle = 45)+
          stat_compare_means(method = "anova",  label.x = 2)+ # Add global anova p-value
          stat_compare_means(label = "p.signif", method = "t.test",
          ref.group = ".all.", hide.ns = TRUE)      # Pairwise comparison against all
```


## 4.3 Question 3: Comparing lapatinib treated breast and cns celllines {#css_id4.3}

As a proven fact, lapatinb crosses the brain-blood barrier. Since breast cancer tends to spread and form metastasis in brain cells and lapatinib is used in anti-breast cancer therapy, we wanted to analyse our data for similar effects after lapatinib treatment in breast and cns tissue cells. 

At first, we selected lapatinib response genes by creating matrices with cns- and breast-cell line fold change values. 
```{r, echo=FALSE}
L_fc <- select(Fold_Change, contains("Lapa"))
L_fc <- as.data.frame(t(L_fc))
rownames(Metadata) <- Metadata$sample


L_treated <- select(Treated, contains("Lapa"))
L_treated <- t(L_treated)
L_untreated <- select(Untreated, contains("Lapa"))
L_untreated <- t(L_untreated)



# selecting breast Lapatinib samples
breast <- Metadata[Metadata[,'tissue']=="Breast",]
rownames(breast) <- breast$sample
rownames(breast) <- gsub(x = rownames(breast), pattern = "-", replacement = ".")  

breastFC <- subset(L_fc, rownames(L_fc) %in% rownames(breast))
breastTreated <- subset(L_treated, rownames(L_treated) %in% rownames(breast))
breastUntreated <- subset(L_untreated, rownames(L_untreated) %in% rownames(breast))#


# selecting CNS Lapatinib samples
cns <- Metadata[Metadata[,'tissue']=="CNS",]
rownames(cns) <- cns$sample
rownames(cns) <- gsub(x = rownames(cns), pattern = "-", replacement = ".")

cnsFC <- subset(L_fc, rownames(L_fc) %in% rownames(cns))
cnsTreated <- subset(L_treated, rownames(L_treated) %in% rownames(cns))
cnsUntreated <- subset(L_untreated, rownames(L_untreated) %in% rownames(cns))
```

Next, we performed a paired t-test of treated and untreated cns and breast samples, respectively, to determine statistical significant fold change values. Since we are performing multiple testing over different cell lines from breast and cns tissues, we want to adjust our p value by performing a Benjamini Hochberg correction.

```{r}
#performing a paired t-test of treated and untreated samples
t_test_cns <- col_t_paired(cnsTreated, cnsUntreated, alternative = "two.sided", mu = 0,conf.level = 0.95)
t_test_breast <- col_t_paired(breastTreated, breastUntreated, alternative = "two.sided", mu = 0,conf.level = 0.95)

#obtaining Benjamini-Hochberg adjusted p-values
pval_cns <- t_test_cns$pvalue
pval_breast <- t_test_breast$pvalue

fdr_cns <- p.adjust(pval_cns, "BH")
fdr_breast <- p.adjust(pval_breast, "BH")


#obtaining mean FC values over all samples 
breastFCm <- as.numeric(colMeans(breastFC))
cnsFCm <- as.numeric(colMeans(cnsFC))
genes <- colnames(breastFC)
```


### 4.3.x Volcano plot {#css_class4.3.x}

For visualization of the biological significance (fold change values) and the statistical significance (adjusted p-values), we plotted our data in an interactive volcano plot. 

```{r}
## breast volcano plot
#creating a matrix containg all needed values for plotting
diff_df_breast <- data.frame(gene = genes, Fold = breastFCm, FDR = fdr_breast)
diff_df_breast$absFold <- abs(diff_df_breast$Fold)
head(diff_df_breast)


# add a grouping column; default value is "not significant"
diff_df_breast$group <- "NotSignificant"



# change the grouping for the entries with significance but not a large enough Fold change
diff_df_breast[which(diff_df_breast['FDR'] < 0.5 & (diff_df_breast['absFold']) < 0.2 ),"group"] <- "Significant"

# change the grouping for the entries a large enough Fold change but not a low enough p value
diff_df_breast[which(diff_df_breast['FDR'] > 0.5 & (diff_df_breast['absFold']) > 0.2 ),"group"] <- "FoldChange"

# change the grouping for the entries with both significance and large enough fold change
diff_df_breast[which(diff_df_breast['FDR'] < 0.5 & (diff_df_breast['absFold']) > 0.2 ),"group"] <- "Significant&FoldChange"


# Find and label the top peaks.
top_peaks_breast <- diff_df_breast[with(diff_df_breast, order(Fold, FDR)),][1:10,]
top_peaks_breast <- rbind(top_peaks_breast, diff_df_breast[with(diff_df_breast, order(-Fold, FDR)),][1:10,])


# Add gene labels for all of the top genes we found
# creating an empty list, and filling it with entries for each row in the dataframe
# each list entry is another list with named items that will be used 
a <- list()
for (i in seq_len(nrow(top_peaks_breast))) {
  m <- top_peaks_breast[i, ]
  a[[i]] <- list(
    x = m[["Fold"]],
    y = -log10(m[["FDR"]]),
    text = m[["gene"]],
    xref = "x",
    yref = "y",
    showarrow = TRUE,
    arrowhead = 0.5,
    ax = 20,
    ay = -40
  )
}

plot_breast <- plot_ly(data = diff_df_breast, x = diff_df_breast$Fold, y = -log10(diff_df_breast$FDR), type = "scatter", text = diff_df_breast$gene, mode = "markers", color = diff_df_breast$group) %>% 
  layout(title ="Volcano Plot of Lapatinib breast cancer samples", 
         xaxis = list(title="log2 Fold Change"),
         yaxis = list(title="FDR")) %>%
  layout(annotations = a)
plot_breast

###thresholds still need to be discussed
```
```{r}

## CNS volcano plot

diff_df_cns <- data.frame(gene = genes, Fold = cnsFCm, FDR = fdr_cns)
diff_df_cns$absFold <- abs(diff_df_cns$Fold)
head(diff_df_cns)


# add a grouping column; default value is "not significant"
diff_df_cns$group <- "NotSignificant"


# change the grouping for the entries with significance but not a large enough Fold change
diff_df_cns[which(diff_df_cns['FDR'] < 0.5 & (diff_df_cns['absFold']) < 0.2 ),"group"] <- "Significant"

# change the grouping for the entries a large enough Fold change but not a low enough p value
diff_df_cns[which(diff_df_cns['FDR'] > 0.5 & (diff_df_cns['absFold']) > 0.2 ),"group"] <- "FoldChange"

# change the grouping for the entries with both significance and large enough fold change
diff_df_cns[which(diff_df_cns['FDR'] < 0.5 & (diff_df_cns['absFold']) > 0.2 ),"group"] <- "Significant&FoldChange"


# Find and label the top peaks..
top_peaks_cns <- diff_df_cns[with(diff_df_cns, order(Fold, FDR)),][1:10,]
top_peaks_cns <- rbind(top_peaks_cns, diff_df_cns[with(diff_df_cns, order(-Fold, FDR)),][1:10,])


a <- list()
for (i in seq_len(nrow(top_peaks_cns))) {
  m <- top_peaks_cns[i, ]
  a[[i]] <- list(
    x = m[["Fold"]],
    y = -log10(m[["FDR"]]),
    text = m[["gene"]],
    xref = "x",
    yref = "y",
    showarrow = TRUE,
    arrowhead = 0.5,
    ax = 20,
    ay = -40
  )
}

plot_cns <- plot_ly(data = diff_df_cns, x = diff_df_cns$Fold, y = -log10(diff_df_cns$FDR),type = "scatter", text = diff_df_cns$gene, mode = "markers", color = diff_df_cns$group) %>% 
  layout(title ="Volcano Plot of Lapatinib CNS cancer samples",
         xaxis = list(title="log2 Fold Change"),
         yaxis = list(title="FDR"))%>%
  layout(annotations = a)
plot_cns

```

To visualize the expression patterns of the top peak genes present in both tissue types we calculated a heat map, comparing the expression profiles of the two tissue types.

```{r}
# selecet top peak genes common in cns and breast tissue
tpb_comparison <- subset(top_peaks_breast, gene %in% top_peaks_cns$gene)
tpc_comparison <- subset(top_peaks_cns, gene %in% top_peaks_breast$gene)


# order common genes alphabetically
tpb_comparison <- tpb_comparison[order(tpb_comparison$gene),]
tpc_comparison <- tpc_comparison[order(tpc_comparison$gene),]


## creating heat map of FCs to compare values 
cor_mat <- cbind("breast" = tpb_comparison$Fold, "cns" = tpc_comparison$Fold)
rownames(cor_mat) <- tpb_comparison$gene
data <- read.delim


pheatmap(
  mat               = cor_mat,
  color             = magma(10),
  border_color      = "black",
  show_colnames     = TRUE,
  show_rownames     = TRUE,
  drop_levels       = TRUE,
  fontsize          = 14,
  main              = "Comparison:
  FC levels of cns and breast top peak genes"
)
```

```{r, echo=FALSE}
#correlation between top peak gene expression patterns in breast and cns tissues treated by lapatinib
cor_mat <- as.data.frame(cor_mat)
dif.FC.BC = data.frame(breast = cor_mat$breast -  mean(t(breastFC)), cns = cor_mat$cns -  mean(t(cnsFC))) #breast data - mean value, cns data - mean value
dif.FC.BC$gene = rownames(cor_mat)
```

```{r, fig.height = 7, fig.width = 14}
# PLot

b = ggplot(dif.FC.BC,aes(x = gene, y = breast)) +
  geom_bar(stat = "identity", fill = "skyblue") + 
  geom_text(aes(label = round(breast, 2)), vjust = -0.5, color = "black", size = 3) + 
  coord_flip() + labs(title = "Mean graph plot of Fold Change for breast top peak genes")


c = ggplot(dif.FC.BC,aes(x = gene, y = cns)) +
  geom_bar(stat = "identity", fill = "skyblue") + 
  geom_text(aes(label = round(cns, 2)), vjust = -0.5, color = "black", size = 3) + 
  coord_flip() + labs(title = "Mean graph plot of Fold Change for CNS top peak genes")

ggarrange(b, c,
          labels = c("A", "B"),
          ncol = 2, nrow = 1)
rm(b,c)
```

At last, we investigated the expression regulation in both tissue types by calculating the pearson correlation between the fold change values of the top peak genes.

```{r}
#correlation
cor(cor_mat$breast, cor_mat$cns, method = "pearson")
```

The obtained correlation value of 0.92 indicates that lapatinib treatment yields a similar effect on gene expression in breast and cns tissue cells and therefore might be possible to treat cancer in brain metastasis.  
</div>
